Switch equivalence for left-linear rewriting systems

### 2024-01-31 icalp/fusione.tex
summary

Graph-like structures are used in uncountably many setting for representing the dynamics of software systems: the graph represents the entities in the states and their relations, while rewriting steps, modifying the graph, model computational steps which produce state changes.

Rewriting over graph-like structures is elegantly captured by
DPO-rewriting over adhesive categories and their
variants~\cite{lack2005adhesive,ehrig2006weak}.  This has been largely
studied for so-called linear rewriting rules. Roughly speaking, with
linear rules a rewriting step can express the fact that some entities
in the current state are removed, some other are needed for the
rewriting step to occur but preserved and some new entities are
generated.

The computation described by a rewriting system can be naturally
interpreted as a distributed computation and a notion of independence
between rewriting steps naturally emerges. When different rules are
applied at ``distinct'' parts of the state, possibly overlapping on
preserved items, the rewriting steps can be naturally deemed as
independent and their application can occur in any order and possibly
concurrently. This phenomenon is formally captured by the so-called
Church-Rosser theorem of DPO rewriting and shift equivalence over derivations~\cite{CMREHL:AAGT}. This represent the basis for
viewing sequences of rewriting in the DPO approach as concurrent
computations and derive corresponding concurrent semantics. Derivation
traces can be quotiented by an equivalence relation which identifies
sequences of rewritings which differs only for the order of
independent steps, leading to so-called derivation traces. In turn,
the dependencies this setting can be captured by a classical structure
in concurrency theory, the so-called prime event structures~\cite{NPW:PNES} where
dependencies are reduced to causality and conflict~\cite{Handbook,Bal:PhD,Sch:RRSG}.

In some situations linear rules are not sufficiently expressive for capturing the dynamic of a system. This is the case when  a computational step is required not only to consume, preserve and generated, but also ``merge'' parts of the state. This happens, e.g.,
in nominal calculi where, as a result of name passing,
the received name is identified with a local one at the
receiver~\cite{CVY:ESSPE,Gad07} or in the modelling of bonding in biological/chemical processes~\cite{PUY:MBPE}.
 add EGGS, something about string diagrams?

Technically, describing fusions, in the DPO jargon, requires to move from linear to left-linear rewriting rules. While left-linear rules have been considered in various instances (mainly for graph rewriting( and some results concerning the corresponding rewriting theory have been put forward, to the best of our knowledge the phenomena arising when rewriting with left-linear rules in the general setting of adhesive categories (which capture most ``applications'') have not been systematically explored.

Here we perform this exploration by observing that a number of basic
properties of rewriting which play a role in the corresponding
rewriting and concurrency theory need a deep rethinking for
left-linear rules and we propose solutions for the problems which
arise.

Firstly, Church-Rosser theorem does not work out of the box: the
independence condition for adhesive categories does not extend
smoothly to left-linear rules, i.e., existence of an independence pair
does not ensure the possibility of switching two derivation and a
refined notion, involving the concept of a so-called filler has to be
introduced. The problem was identified already
in~\cite{baldan2011adhesivity} where a suitable subclass of adhesive
categories is identified where independence as induced by a filler
coincide with independence as induced by an independence pair.

Having a notion of independence between computation steps sits at the core of the possibility of viewing an execution trace as a concurrent computation. This is obtained by considering traces up to an equivalence that identifies traces differing only for the order of independent steps (e.g., interchange law~\cite{Mes92} in term rewriting, shift
equivalence~\cite{CMREHL:AAGT} in graph rewriting or permutation
equivalence~\cite{JJL80} in $\lambda$-calculus). While for linear rewriting systems this works smoothly, for left-linear ones one has to face the problem that if two rewriting steps $r_a r_b$ are independent, if we switch $r_a$ after $r_a$ and then switch it back we can obtain a computation which is not the original one, making the theory of derivation traces inconsistent. In order to solve this problem, we work in a subclass of adhesive categories where derivation steps can be switched in an ``essentially unique way''. We also identify a large class of adhesive categories, arising as presheaves, where the condition is automatically satisfied. [this solves also the previous problem, right?]

Additionally, a central property for linear rewriting systems is the fact that independence of two rewriting steps is a global notion, independently of the context in which the rewritings take place. This is at the basis of the possibility of capturing the dependencies between rewriting steps in terms of two basic relations, causality and conflict, investigated in~~\cite{Handbook,Bal:PhD,Sch:RRSG}.

This fails for left-linear rules where, as observed
in~\cite{baldan2017domains}, for graph rewriting, the possibility of
merging parts of the state leads to a sort of disjunctive causality,
i.e., a rewriting step, say $r_c$ can be enabled by two different
rewriting steps $r_a$ and $r_b$, in a way that in a derivation
$r_a r_b r_c$ steps $r_b$ and $r_c$ are independent since the presence
of $r_a$ in the past is sufficient for enabling $r_a$, while in
$r_b r_c r_a$ , the steps $r_b$ and $r_c$ are not independent. The
paper shows that for left-linear rewriting systems a special form of
globality of independence can be recovered.

The concurrent semantics of linear graph rewriting systems over
adhesive categories can be captured by prime event structures and
prime algebraic domains. The idea consists in considering a partially
ordered domain, where elements are concurrent computations
(derivations up to shift equivalence), and the order relates two
computations when the latter is an extension of the former.
%
Events are then identified with computations consisting of a rewriting
step with all its causes. This is generalised to adhesive categories
in~\cite{heindel2009category}. [IS THIS IN TOBIAS?]

In~\cite{baldan2017domains} it is observed for graph rewriting systems that the presence of fusions, the same event can be enabled by different minimal sets of events, i.e., a central property of the domain of configuration, namely stability, fails. This prevents the identification of a proper notion of causality and makes prime event structures and prime domains inadequate to represent computations with fusions. It is argued that the right class of structures capturing the concurrent behaviour in the presence of fusion are connected event structures and weak prime algebraic domains.
%
An event is still a computation that cannot be decomposed
as the join of other computations, hence, in order theoretical terms,
it is an irreducible.
%
However, due to unstability, irreducibles are not necessarily primes:
two different irreducibles can represent the same computation step
with different minimal enablings.  In~\cite{baldan2017domains} this
is worked out for DPO rewriting systems over graphs. However the proof
relies on a wrong characterisation of independence which fails for non
linear rules. Here we rework in depth the result and extend it to DPO
rewriting over a large class of adhesive categories.

>>>>

Rewriting systems over graphs are used in uncountably many setting for
representing the dynamics of software systems: the graph represents
the entities in the states and their relations, while rewriting steps,
modifying the graph, model computational steps which produce state
changes.

Rewriting over many kind of graphs is elegantly captured by the
so-called \emph{double pushout} (DPO) approach to rewriting. A
convenient fact is that DPO rewriting can be defined over adhesive
categories and their
variants~\cite{lack2005adhesive,ehrig2006weak}. Working in the
framework of adhesive categories allows one to devise the foundations
of a rewriting theory which can then be instantiated to the setting of
interest, avoiding to prove over and over similar ad-hoc results in
each specific context.

Rewriting over adhesive categories has been widely studied in the last
two decades and most of the research focused on the so-called linear
rewriting rules. Roughly speaking, linear rewriting rules can specify
that, in a rewriting step, some entities in the current state are
removed, some other are needed for the rewriting step to occur but
remain unaltered and some new entities can be generated.

The computation described by a rewriting system can be naturally
interpreted as a distributed computation and a notion of independence
between rewriting steps naturally emerges. When different rules are
applied at ``distinct'' parts of the state, possibly overlapping on
preserved items, the rewriting steps can be naturally deemed as
independent and their application can occur in any order and possibly
concurrently. This phenomenon is formally captured by the so-called
Church-Rosser theorem of DPO rewriting~\cite{CMREHL:AAGT}.
%
Having a notion of independence between computation steps sits at the
core of the possibility of viewing an execution trace as a concurrent
computation. This is obtained by considering traces up to an
equivalence that identifies traces differing only for the order of
independent steps, referred to as shift equivalence~\cite{CMREHL:AAGT}
(similar to interchange law~\cite{Mes92} in term rewriting or
permutation equivalence~\cite{JJL80} in the $\lambda$-calculus).

In some situations rewriting with linear rules are not sufficiently
expressive for capturing the dynamic of a system. This is the case
when a computational step is required not only to consume, preserve
and generated, but also ``merge'' parts of the state. This happens,
e.g., in nominal calculi where, as a result of name passing, the
received name is identified with a local one at the
receiver~\cite{CVY:ESSPE,Gad07} or in the modelling of bonding in
biological/chemical processes~\cite{PUY:MBPE}. A similar mechanisms is
at the core of e-graphs (equality graphs), recently used for
rewrite-driven compiler optimizations~\cite{WNW:egg}. Rather than
explicitly merging nodes corresponding to equal expression, the idea
here - conceptually and, as we will see, also mathematically similar -
is to maintain an equivalence\todo{P: more precisely a congruence}
over the nodes.  \todo{P: la sequenza di lavori su questo \`e ampia,
  vedere se c'\`e qualcosa d'altro}

\todo{P: something about string diagrams?}

Technically, to acquire the expressiveness needed for expressing
fusions, i.e., the merge of some elements in the state, requires to
move from linear to left-linear rewriting rules. While left-linear
rules have been considered in various instances (mainly in categories
of graphs) and some results concerning the corresponding rewriting
theory have been put forward, to the best of our knowledge the
phenomena arising when rewriting with left-linear rules in the general
setting of adhesive and quasi-adhesive categories (which capture most
``applications'') have not been systematically explored. ADD REFS

Here we perform this exploration by observing that for left-linear
rules a number of fundamental properties which play a role in the
corresponding theory of rewriting need a deep rethinking.

Firstly, Church-Rosser theorem does not work out of the box. In
particular, sequential independence between rewriting steps~\cite{}
expressed in terms of the existence of a so-called independence pair
is a canonical and well-understood notion of independence in the
linear case. \todo{P: Too technical? Forse qui e' piu' esplicativo un
  esempio} Whenever two steps are sequential independent
\begin{itemize}
\item they can be switched and 
\item the result of the switch is uniquely determined.
\end{itemize}

Here we observe that, in general, both property can fail: sequential
independent steps could be not switchable and different independent
pairs could exist leading to different switches of the same rewriting
steps.  A refined notion of independence can be considered to ensure
that independent rewriting steps can be switched, based in the concept
of a so-called filler. The problem was identified already
in~\cite{baldan2011adhesivity} where a suitable subclass of adhesive
categories is identified where independence as induced by a filler
coincide with sequential independence (as induced by an independence
pair). Here this is extended to $\mathcal{M}$-adhesive categories.

We face the problem in the setting of $\mathcal{M}$-adhesive
categories (which capture adhesive and quasi-adhesive categories).
The natural idea, put forward in~\cite{baldan2017domains}, of limiting
sequential independence to those cases in which the switch is possible
and uniquely determined falls short.

Rather we focus on a class of rewriting systems, the so-called
\emph{very tame rewriting systems} in $\mathcal{M}$-adhesive
categories where sequential independence ensures existence and
uniqueness of the switch and argue that this is the right setting for
dealing with left-linear rewriting systems.

On the one hand we argue that very tame rewriting systems capture most
categories of interest for rewriting. In particular, we single out a
class of left-linear rewriting systems on graph-like structures which
are very tame and we show that rewriting systems on (term)graphs with
equivalences resempling e-graphs are also very tame. As a sanity
check, also linear rewriting systems on $\mathcal{M}$-adhesive
categories are very tame.

On the other hand we show that an appropriate theory of rewriting can
be developed for very tame rewriting systems, recovering, sometimes in
weakened form, results which hold in the linear case.  Local
Curch-Rosser is fully recovered for very tame rewriting systems.

For linear rewriting systems sequential independence of derivations
can be also characterised in terms of the existence of a suitable
isomorphism between the colimit of the derivations, the so-called
consistent permutations (which are also strictly related to the notion
of process). This result, which is folklore for categories of graphs,
is explicitly proved here for $\mathcal{M}$-adhesive categories. Based on such result one can deduce two fundamental consequences:
\begin{enumerate}
\item when swicthing a derivation consisting of several steps, the result only depends on the overall permutation, not on the order of the swicthes
\item independence of rewriting steps is a ``global property'', independent of the position in which these steps are in the derivation.
\end{enumerate}

We show that for left-linear rewriting systems the characterisation of
shift equivalence based on consistent permutations fails: the
existence of a consistent permutation is only necessary but not
sufficient, hence it cannot be used for proving (1) and (2). However property (1) can be still proved to holds. Instead only a weak form of (2) can be proved. This is intimately related to the possibility of expressing disnjuctive forms of causality.

\todo[inline]{
This fails for left-linear rules where, as observed
in~\cite{baldan2017domains}, for graph rewriting, the possibility of
merging parts of the state leads to a sort of disjunctive causality,
i.e., a rewriting step, say $r_c$ can be enabled by two different
rewriting steps $r_a$ and $r_b$, in a way that in a derivation
$r_a r_b r_c$ steps $r_b$ and $r_c$ are independent since the presence
of $r_a$ in the past is sufficient for enabling $r_a$, while in
$r_b r_c r_a$ , the steps $r_b$ and $r_c$ are not independent. }
